Debugging of RxJS-based Applications

### 2020-08-20 paper.tex
Restructure Intro

\documentclass[sigplan,screen,review]{acmart}
\usepackage{ccicons}
\usepackage{hyperref}
\usepackage{minted}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{./tikzit}

@@

% Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
\acmConference[SPLASH '20]{SPLASH '20: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
\title{Debugging of RxJS-based Applications}
\author{Manuel Alabor}
\affiliation{
	\institution{HSR University of Applied Sciences}
	\city{Rapperswil}
	\country{Switzerland}
}
\email{malabor@hsr.ch}

@@

According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.

@@

These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry. Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities. For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when applying RP. Based on this collected evidence, we will validate their statements in an observational study using RxJS and search for an answer to our first research question:

@@

We will conclude this introductory section with the clarification of important terms. Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports. We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RP code in Section~\ref{sec:discussion}. Finally, we will answer RQ2 in Section~\ref{sec:future} ``Future Work''.

@@

\begin{figure}[H]
	\centering
	\tikzfig{debugging-process-model}
	\Description{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
	\label{fig:debugging-process-model}
\end{figure}

@@

@@

On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies. E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs. In the end, we were able to gather compelling statements which we are going to present in this section.

@@

After we built a first intuition for how software engineers work with RP by evaluating the interviews, we were interested in more RxJS-specific, hands-on experiences. We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request. After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another set of reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.

@@

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our interview partners. R5 described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information at any time. After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed numbers kept changing where we did not expected them to. In other places, they were not rerendered where they were supposed to after, e.g. after we changed them in the system,'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several times. Other parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future. (We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion}.)

@@

\section{Validation of Interviews and War Stories}
\label{sec:study}

@@

Before we elaborate on a solution in order to overcome the challenges identified with RQ1, we are going to present a selection of RP specific debugging utilities available today.

\section{Discussion of Previous Efforts}
\label{sec:discussion}

\subsection{rxjs-spy}

\emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables. Once an observable is tagged with an arbitrary string identifier, the library can generate trace logs when a value is emitted as well as when individual lifecycle events occur (i.e. subscribe, unsubscribe, complete and error). Ideally, tagged Observables are created during development when the dataflows are composed for the first time. Like the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-debugging}, the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-spy-tag} on Line~3 is completely transparent to the actual dataflow.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		import { create } from 'rxjs-spy';
		import { tag } from 'rxjs-spy/operators';

		const spy = create(); // Create monitor
		spy.log(/multiply/);  // Log tags matching
		                      // RegEx /multiply/

		interval(1000).pipe(
		  map(i => i * 2),
		  tag('multiply'),    // Tag with "multiply"
		  map(i => i - 1),
		  tag('subtract')     // Tag with "subtract"
		  take(2)
		).subscribe();
	\end{minted}
	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~10 and 12.}
	\label{lst:rxjs-spy-tag}
\end{listing}

The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log}.

\begin{listing}[H]
	\begin{minted}{Typescript}
		Tag = multiply; notification = subscribe
		Tag = multiply; notification = next; value = 0
		Tag = multiply; notification = next; value = 2
		Tag = multiply; notification = unsubscribe
	\end{minted}
	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
	\label{lst:rxjs-spy-log}
\end{listing}

Additional features are available through the debuggers console interface. E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately. The engineer can then emitted these values one after another manually or resume all of them at once.

\subsection{rxfiddle}

\subsection{RxViz}

\subsection{Reactive Inspector}

Along with the RP specific debugging features, Reactive Inspector experiments with the omniscient debugging concept: The time travel functionality allows to record dataflows. These recordings can then be replayed and inspected at a later time without actual re-execution of the program.

\subsection{Comparison}

Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools. Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy). They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector). Whereas Reactive Inspector for REScala provides an ``all-in-one'' solution with reactive breakpoints, omniscient debugging capabilities, and full IDE integration, the RxJS-related tools provide smaller, differentiated feature sets in contrast. Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or call for manual code preparation or modification (rxjs-spy).

\begin{table}[H]
	\caption{Feature Matrix for Discussed RP Debugging Tools}
	\begin{tabular}{@{}lllll@{}}
	\toprule
					   & \textbf{UI} & \textbf{Integration}                                               & \textbf{Flow Control}                                              & \textbf{RP Runtime} \\ \midrule
	Reactive Inspector & Visual      & Eclipse IDE                                                        & Full                                                               & REScala             \\ \midrule
	rxfiddle           & Visual      & Web App                                                            & None       & RxJS                \\ \midrule
	rxjs-spy           & Textual     & \begin{tabular}[c]{@{}l@{}}Browser Console\\and Code\end{tabular} & Limited & RxJS                \\ \midrule
	Rx Visualizer      & Visual      & Web App                                                            & None       & RxJS                \\ \bottomrule
	\end{tabular}
	\label{tab:compare-tools}
\end{table}

@@

@@

\begin{acks}
	We would like to
\end{acks}

>>>>

\documentclass[sigplan,screen,review]{acmart}
% Class options:
%  - review: Show line numbers for reviewers
%  - nonacm: Hide ACM conference stuff

\usepackage{hyperref}
\usepackage{minted}
\usepackage{booktabs}
\usepackage{./tikzit}

@@

% Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
\acmConference[SPLASH '20]{SPLASH 2020: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
\title{Debugging of RxJS-based Applications}
\author{Manuel Alabor}
\affiliation{
	\institution{HSR University of Applied Sciences}
	\city{Rapperswil}
	\country{Switzerland}
}
\email{malabor@hsr.ch}

@@

According to the IEEE Standard Glossary of Software Engineering, \emph{debugging} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.

@@

These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow-specific inquiry. Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities. For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when applying RP. Based on this collected evidence, we will validate their statements in an observational study using RxJS and search for an answer to our first research question:

@@

We will conclude this introductory section with the clarification of important terms and a view on known RP debugging utilities. Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports. We present our observational study intended to validate former insights in Section~\ref{sec:study} which allow us to answer RQ1. Before our final conclusion, we will answer RQ2 in Section~\ref{sec:future} ``Future Work'' and review the threats to validity regarding our study in Section~\ref{sec:threats}.

@@

\begin{figure}
	\centering
	\tikzfig{debugging-process-model}
	\Description{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
	\label{fig:debugging-process-model}
\end{figure}

@@

\subsection{RxJS Debugging Utilities}

\subsubsection{rxjs-spy}

\emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables. Once an observable is tagged with an arbitrary string identifier, a monitor can generate trace logs whenever a value is emitted as well as when individual lifecycle events occur (i.e. subscribe, unsubscribe, complete and error). Ideally, tagged Observables are created during development when the dataflows are composed for the first time. Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an Observable emits a value. It cannot modify or influence the emitted value nor the Observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual dataflow.

\begin{listing}
	\begin{minted}{TypeScript}
		import { create } from 'rxjs-spy';
		import { tag } from 'rxjs-spy/operators';

		const spy = create(); // Create monitor
		spy.log(/multiply/);  // Log tags matching
		                      // RegEx /multiply/
		interval(1000).pipe(
		  map(i => i * 2),
		  tag('multiply'),    // Tag with "multiply"
		  map(i => i - 1),
		  tag('subtract')     // Tag with "subtract"
		  take(2)
		).subscribe();
	\end{minted}
	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~9 and 11.}
	\label{lst:rxjs-spy-tag}
\end{listing}

The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log} eventually.

\begin{listing}
	\begin{minted}{Typescript}
		Tag = multiply; notification = subscribe
		Tag = multiply; notification = next; value = 0
		Tag = multiply; notification = next; value = 2
		Tag = multiply; notification = unsubscribe
	\end{minted}
	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
	\label{lst:rxjs-spy-log}
\end{listing}

Additional features are available through the libraries console interface. E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately. The engineer can then emitted these values one after another manually or resume all of them at once.

\subsubsection{rxfiddle}

\subsubsection{RxViz}

@@

On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies. E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs. In the end, we were able to conduct interviews with 10 engineers and collect war stories from another five.

@@

After we built an intuition for how software engineers work with RP by evaluating the interviews, we were interested in more RxJS-specific, hands-on experiences. We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request. After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another set of reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.

@@

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our interview partners. R5 described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information at any time. After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed numbers kept changing where we did not expected them to. In other places, they were not rerendered where they were supposed to after, e.g. after we changed them in the system,'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several times. Other parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.

@@

\section{Validation}
\label{sec:study}

@@

@@

\textbf{TODO Refer to Reactive Inspector}

\subsubsection{Reactive Inspector}

Along with the RP specific debugging features, Reactive Inspector contains  omniscient debugging concepts as well: The time travel functionality allows to record dataflows. These recordings can be replayed and inspected at a later time without an actual re-execution of the recorded program.

@@

\begin{acks}

\end{acks}
